---
title: "Man the Hunter, Natural Language Processing"
author: "Matthew L. Sisk, Phd"
date: "4/13/2026"
format: html
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 10
    fig.height: 5.625
---

## Loading data

This just loads the relevant packages for text analysis and downloads the text of the book from the internet archive.

```{r}
#| warning: false
#| message: false


rm(list = ls())  

library(tidytext)
library(textstem)
library(tidyverse)
library(wordcloud)
library(RColorBrewer)

if (!file.exists("MtH.txt")){
  download.file("https://ia801705.us.archive.org/30/items/ManTheHunter/Man%20the%20Hunter_djvu.txt",
                destfile = "MtH.txt")
  
}
mth_unmodified <-readLines("MtH.txt")



```

## Processing the text

This is cleaning up the text. Mainly removing the line breaks and rejoining words that were split between lines with a hyphen, taking out the page numbers and making all of the special characters (like apostrophes) consistent

```{r}


preface_starts <- 75
part1_starts <- 746
references_starts <- 36991


raw <- mth_unmodified[part1_starts:references_starts] 


modified_text <- paste(raw, collapse = "\n") %>%
  gsub("['\u2018\u2019\u201A\u201B\u0060\u00B4]", "'", .) %>% #standardize apostrophes
  gsub("\r\n", "\n", .) %>% #fix line endings
  gsub("([a-z])-\\s*\n\\s*([a-z])", "\\1\\2", .) %>% #remove hyphenated lines
  gsub("(?m)^\\s*\\d+\\s*$", "", ., perl = TRUE) %>%# cut page numbers
  gsub("huntergatherers?\\b", "hunter gatherer", .) %>%
  gsub("huntinggathering", "hunting gatherering", .)
 

# text
```

## Tolkenizing

After cleaning, the text is usually processed a little more. One process is called lemmantization, or making words into their roots (like cats becoming cat or running becoming run). In this context, it does a pretty good job, leaving both hunt (the verb and hunter (the noun) as separate.

The second step is removing stop words: common words that are unrelated to the sentiment or topic. This could be things like "is, he, or, and" These tend to be standard lists, sometimes with additional ones added. The package I use here has three different ones that you can either use together or individually. The problem is one in particular (an older one called onix, that was meant for information retrieval and searching) contains the words "man" and "men". It does not contain "woman" or "women" The reason for this is presumably that man/men was thought to be shorthand for "human" so if you are searching a text for "man's evolution" the word "man" is not important to the actual search and could lead you astray. Kind of classic gender bias in language and text. The thing in this context though is that it should have been really obvious since its the title of the book. And I should note, that I did a quick scan of the main stop word lists and onix is the only one that has man/men on it.

So, in this part, I am making the stop word list in three ways. 1: just using one of the main packages for language processing as is. One, taking that list and removing "man" and "men" from it but otherwise leaving it as is. And last, just not using the onix stopword list.

```{r}
base_stopwords <- stop_words 
cut_stopwords <- stop_words |>
  filter(!(word %in% c("man", "men")))
no_onix_stopwords <- stop_words |> filter(lexicon != "onix")


bag_of_words <- tibble(text = modified_text) %>%
  unnest_tokens(word_orig, text) %>%
  mutate(word_orig = gsub("'s$", "", word_orig)) %>% #removing possessives
  filter(nchar(word_orig) > 2) %>%
  mutate(word = lemmatize_words(word_orig)) %>%
  filter(!str_detect(word, "[0-9]")) #removing things that are just numbers

full_sum_table <- bag_of_words %>%
  count(word, sort = T)%>%
  mutate(orig = word %in%  base_stopwords$word,
         no_onix = word %in% no_onix_stopwords$word,
         man_back = word %in% cut_stopwords$word)%>%
  filter(!orig | !no_onix | !man_back)



```

## Word Clouds

Here i just made a way to do the word cloud so it kind of looks like the one in the paper

```{r}


make_word_cloud <- function(table, stopword_col, seed = 44,
                            width = 1600, height = 900, 
                            file = NULL) {
  top200 <- table %>%
    filter(!{{ stopword_col }}) %>%
    slice_head(n = 200)
    par(mar = c(0, 0, 0, 0))          # zero margins

  if (!is.null(file)) {
    png(file, width = width, height = height, res = 150)
  }
  
  set.seed(seed)
  wordcloud(
    words        = top200$word,
    freq         = top200$n,
    min.freq     = 1,
    max.words    = 200,
    random.order = FALSE,
    rot.per      = 0.0,
    colors       = "black",
    scale        = c(2, 0.5),
    random.color = FALSE
  )
  
  if (!is.null(file)) dev.off()
}


```

```{r}
#| fig-width: 10
#| fig-height: 5.625
make_word_cloud(table = full_sum_table, stopword_col = orig)

```

For the original data (above), it looks basically like what came before

```{r}
#| fig-width: 10
#| fig-height: 5.625
make_word_cloud(table = full_sum_table, stopword_col = man_back)

```

If we just add man back in, it looks a little better. "Man" is mentioned around 500 times and is the 4th most common word

```{r}
#| fig-width: 10
#| fig-height: 5.625
make_word_cloud(table = full_sum_table, stopword_col = no_onix)
```

If we just skip the onix stop word list entirely it gets even a little more interesting. One of the other main words that was removed in their analysis was "group" which is actually the most common word, and kind of relevant in this context. Here are all the words that were removed from their word cloud, but were mentioned more than 50 times in the original text. Some are fine, but many (e.g. group, man, member, area) contain valuable contextual information for an analysis of human settlement. Using the onix stopword list removed all of that.

```{r}
full_sum_table %>% 
  filter(n >50)%>%
  filter(!no_onix & orig) %>%
  pull(word)
```

## Quick topic classification

To go a little further, we can see if the comparisons between man/women and hunter/gatherer are roughly equal. If both of these topics were both broadly discussed at the conference and in the volume despite the name, then we would expect roughly equal numbers and in a similar variety of contexts. To be clear, some of the processing above and the tokenizing splits the compound word hunter-gatherer ro two separate words (same with hunting-gatherering).

To start, If the only time those words were in there were when hunter-gatherer was mentioned as a single thing, we'd expect them to be roughly equal.

```{r}
# Key words to compare
key_words <- c("man", "men", "woman", "women", 
               "hunt", "hunter", "gather", "gatherer")


full_sum_table %>%
  filter(word %in% key_words) %>%
  mutate(concept = case_when(
    word %in% c("man", "men")           ~ "man/men",
    word %in% c("woman", "women")       ~ "woman/women",
    word %in% c("hunt", "hunter")       ~ "hunt/hunter",
    word %in% c("gather", "gatherer")   ~ "gather/gatherer"
  )) %>%
  group_by(concept) %>%
  summarise(total = sum(n))%>%
  arrange(-total)
```

Hunt/hunting is mentioned about three times as often as gather/gathering (1538 to 568) and man/men is mentioned more than twice as often as woman/women (517 to 220).

But the context that these words are mentioned also plays a role "hunter-gatherer" is a compound word and, we also need to see how many times "gathering" is mentioned independent of that.

```{r}
#Looking at how often gatherer is mentioned without hunter


words_nearby <- bag_of_words %>%
  filter(!word %in% no_onix_stopwords$word) %>%
  mutate(next_word = lead(word),
         pre_word  = lag(word)) %>%
  filter(str_detect(word, "hunt|gath")) %>%
  select(word_orig, pre_word, word, next_word)



# what precedes gatherer?
words_nearby %>%
  filter(str_starts(word, "gath"))%>%
  count(pre_word, word, sort = TRUE) %>%
  head(5)

# what follows hunter?
words_nearby %>%
  filter(str_starts(word, "hunt"))%>%
  count(word, next_word, sort = TRUE) %>%
  head(5)

unique(words_nearby$word)

#to simplify the output, casting all similar words together.
words_nearby <- words_nearby %>%
  mutate(
    word_s      = case_when(
      str_detect(word,      "hunt") ~ "hunt",
      str_detect(word,      "gath") ~ "gather",
      TRUE                           ~ word
    ),
    pre_word_s  = case_when(
      str_detect(pre_word,  "hunt") ~ "hunt",
      str_detect(pre_word,  "gath") ~ "gather",
      TRUE                           ~ pre_word
    ),
    next_word_s = case_when(
      str_detect(next_word, "hunt") ~ "hunt",
      str_detect(next_word, "gath") ~ "gather",
      TRUE                           ~ next_word
    )
  )


tog_gath <- words_nearby %>%
  filter(str_starts(word_s, "gath"))%>%
  count(pre_word_s, word_s, sort = TRUE) %>%
  mutate(percent  = 100*n/sum(n))
tog_gath


# what follows hunter?
tog_hunt_post <- words_nearby %>%
  filter(str_starts(word_s, "hunt"))%>%
  count(word_s, next_word_s, sort = TRUE) %>%
  mutate(percent  = 100*n/sum(n))

tog_hunt_post

# what comes before hunter?

tog_hunt_pre <- words_nearby %>%
  filter(str_starts(word_s, "hunt"))%>%
  count(pre_word_s,word_s,  sort = TRUE) %>%
  mutate(percent  = 100*n/sum(n))

tog_hunt_pre


```

So, \~71 % of the 526 times gathering is mentioned it is immediately preceded by hunter,

but only \~24% of the 1,547 times hunting is mentioned it is not followed by gathering.

And the most common work to come before hunting is (obviously) "man" but its not all that common

Note: If anyone is interested in that one reference to "hunterprovider" with female after it as a possible counter example, i would suggest looking at the original manuscript. Its on page 302
