---
title: "Man the Hunter: A brief Text Analysis"
subtitle: "Examining word frequency in Lee & DeVore (1968)"
author: 
  - name: "Matthew L. Sisk, PhD"
    affiliation:
      - name: "Lucy Family Institute for Data & Society"
      - name: "University of Notre Dame"
date: today
abstract: |
  This document provides a reproducible text analysis of the *Man the Hunter* 
  volume (Lee & DeVore, 1968) in particular to assess a word cloud presented in
  Venkataraman et al. (2026). It demonstrates that the word cloud presented in 
  that paper inadvertently suppressed the word "man" through use of a default 
  stopword list (onix) that treats "man" and "men" as generic terms for "human" 
  while retaining "woman" as a content word. It also examines the relative 
  frequency with which hunting and gathering (and men and women) are discussed 
  in the volume, and the contexts in which these terms appear.
format: 
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    fig-width: 10
    fig-height: 5.625
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false

library(tidytext)
library(textstem)
library(tidyverse)
library(wordcloud)
library(RColorBrewer)
library(DT)

# helper to make scrollable tables
dt_table <- function(df, caption = NULL, pageLength = 10) {
  datatable(df,
            caption    = caption,
            rownames   = FALSE,
            options    = list(pageLength = pageLength,
                              scrollX    = TRUE,
                              dom        = "tip"),
            class      = "compact stripe hover")
}
```

## Data and Methods

The text of *Man the Hunter* (Lee & DeVore, 1968) was obtained from the Internet Archive (plain text format). Analysis was conducted in R using the `tidytext`, `textstem`, and `wordcloud` packages. The table below summarizes the key methodological choices.

```{r methods-table}
tibble(
  Step    = c("Source text", "Lines included", "Lemmatizer",
              "Stopword lists", "Minimum word length"),
  Detail  = c(
    "Internet Archive djvu.txt (Man the Hunter, Lee & DeVore 1968)",
    "Lines 746 to 36,991 (main text only, excluding front matter and references)",
    "textstem R package (dictionary-based lemmatization)",
    "tidytext stop_words: SMART, Snowball, onix",
    "3 characters"
  )
) %>% knitr::kable()
```

## Loading and Processing the Text

The raw text file is downloaded once and cached locally. The following cleaning steps are applied before analysis:

-   Apostrophe variants (curly quotes, backticks) are standardized to a straight apostrophe
-   Windows-style line endings are converted to Unix style
-   Words split across lines with a hyphen are rejoined (e.g., `anthro-\npologists` becomes `anthropologists`)
-   Standalone page numbers are removed
-   The compound `huntergatherer`, created by the line-break joining step, is split back into two tokens

```{r load-data}
if (!file.exists("MtH.txt")) {
  download.file(
    "https://ia801705.us.archive.org/30/items/ManTheHunter/Man%20the%20Hunter_djvu.txt",
    destfile = "MtH.txt"
  )
}

mth_unmodified <- readLines("MtH.txt")

# Line number limits for the main text in the file
preface_starts <- 75        # start of the prefac (not used)
part1_starts      <- 746    # start of main text (after front matter)
references_starts <- 36991  # end of main text (before references)

raw <- mth_unmodified[part1_starts:references_starts]

modified_text <- paste(raw, collapse = "\n") %>%
  gsub("['\u2018\u2019\u201A\u201B\u0060\u00B4]", "'", .) %>%  # standardize apostrophes
  gsub("\r\n", "\n", .) %>%                          # fix line endings
  gsub("([a-z])-\\s*\n\\s*([a-z])", "\\1\\2", .) %>% # rejoin hyphenated linebreaks
  gsub("(?m)^\\s*\\d+\\s*$", "", ., perl = TRUE) %>%  # remove page numbers
  gsub("huntergatherers?\\b", "hunter gatherer", .) %>% # split inaccurate compound
  gsub("huntinggathering", "hunting gathering", .)    # split inaccurate compound
```

## Tokenizing and Lemmatization

Tokenization splits the text into individual words. Lemmatization then reduces each word to its root form; for example, "cats" becomes "cat" and "running" becomes "run". Notably, the lemmatizer correctly distinguishes between *hunt* (the verb) and *hunter* (the noun), keeping them as separate tokens. Possessive suffixes (`'s`) are stripped before lemmatization since the lemmatizer does not handle them consistently.

```{r tokenize}
bag_of_words <- tibble(text = modified_text) %>%
  unnest_tokens(word_orig, text) %>%
  mutate(word_orig = gsub("'s$", "", word_orig)) %>%  # strip possessives
  filter(nchar(word_orig) > 2) %>%                     # minimum 3 characters
  mutate(word = lemmatize_words(word_orig)) %>%
  filter(!str_detect(word, "[0-9]"))      # remove numeric tokens after lemmatizing
                                          # o tavoid e.g."tenth" becoming "10"
```

### Stopword Lists

Stopwords are common words removed before analysis, typically function words like "is", "the", "and". The `tidytext` package bundles three English stopword lists: SMART, Snowball, and onix. When used together (the default), any word appearing in *any* of the three lists is removed.

The onix stopword list is and older and larger list designed for search indexing. As such, it contains "men" and "man" but not "woman" or "women." As the Appendix below shows, this is rare for an English stop word list and is likely the result of its age and original context.

```{r stopwords}
base_stopwords    <- stop_words
cut_stopwords     <- stop_words %>% filter(!(word %in% c("man", "men")))
no_onix_stopwords <- stop_words %>% filter(lexicon != "onix")

tibble(
  `Stopword list`    = c("Default (all three lists)",
                          "man/men restored",
                          "onix excluded entirely"),
  `Contains 'man'`   = c("man" %in% base_stopwords$word,
                          "man" %in% cut_stopwords$word,
                          "man" %in% no_onix_stopwords$word),
  `Contains 'woman'` = c("woman" %in% base_stopwords$word,
                          "woman" %in% cut_stopwords$word,
                          "woman" %in% no_onix_stopwords$word),
  `List size`        = c(nrow(base_stopwords),
                          nrow(cut_stopwords),
                          nrow(no_onix_stopwords))
) %>% knitr::kable()

full_sum_table <- bag_of_words %>%
  count(word, sort = TRUE) %>%
  mutate(
    orig     = word %in% base_stopwords$word,
    no_onix  = word %in% no_onix_stopwords$word,
    man_back = word %in% cut_stopwords$word
  ) %>%
  filter(!orig | !no_onix | !man_back)
```

In this case, using the onix list inadvertently removes a semantically significant word, and part of the title, from the analysis. Three versions of the stopword list are used here for comparison:

1.  Using the default tidytext list of all three.

2.  Using the default list , but manually removing "man" and "men"

3.  Excluding the onix list entirely

## Word Clouds

The word clouds below replicate the style of Figure 3 in Venkataraman et al. (2026), with word size proportional to frequency. Three versions are shown, one for each stopword configuration. Each cloud is followed by a table of the 10 most frequent words in that version.

```{r wordcloud-function}
make_word_cloud <- function(table, stopword_col = NULL, seed = 44,
                            width = 1600, height = 900,
                            file = NULL) {
  if (!missing(stopword_col)) {
    table <- table %>% filter(!{{ stopword_col }})
  }
  
  top200 <- table %>% slice_head(n = 200)  

  if (!is.null(file)) {
    png(file, width = width, height = height, res = 150)
    par(mar = c(0, 0, 0, 0))
  }
  set.seed(seed)
  wordcloud(
    words        = top200$word,
    freq         = top200$n,
    min.freq     = 1,
    max.words    = 200,
    random.order = FALSE,
    rot.per      = 0.0,
    colors       = "black",
    scale        = c(2, 0.5),
    random.color = FALSE
  )
  if (!is.null(file)) dev.off()
}

top10_table <- function(table, stopword_col = NULL) {
  if (!missing(stopword_col)) {        # fixed: was is.null, should be !is.null
    table <- table %>% filter(!{{ stopword_col }})
  }
  
  table %>%
    slice_head(n = 10) %>%
    select(word, n) %>%
    knitr::kable(col.names = c("Word", "Frequency"))
}
```

### 1. Default stopword list

Using the default `stop_words` list, which includes the onix lexicon. "man" and "men" are silently removed; "woman" is retained.

```{r cloud-orig}
#| fig-cap: "Word cloud with default stopword list applied; 'man' and 'men' are absent."
make_word_cloud(table = full_sum_table, stopword_col = orig)
```

```{r hunt-count}
#| include: false
hunt_count_orig <- full_sum_table %>%
  filter(!orig, word == "hunt") %>%
  pull(n)
```

The 10 most frequent words using the default stopword list are shown below. Venkataraman et al. (2026) report approximately 835 occurrences of "hunt" in their Figure 3; our replication yields `r hunt_count_orig` occurrences. The small discrepancy likely reflects minor differences in text extraction or preprocessing between the two analyses.

```{r top10-orig}
top10_table(full_sum_table, orig)
```

### 2. With "man" and "men" restored

Identical to above, but with "man" and "men" removed from the stopword list (so included in the final product).

```{r cloud-manback}
#| fig-cap: "Word cloud with 'man' and 'men' restored. 'Man' now appears prominently, consistent with its frequency in the text."
make_word_cloud(table = full_sum_table, stopword_col = man_back)
```

```{r man-rank}
#| include: false
man_count <- full_sum_table %>% filter(word == "man") %>% pull(n)
man_rank  <- full_sum_table %>%
  filter(!man_back) %>%
  mutate(rank = row_number()) %>%
  filter(word == "man") %>%
  pull(rank)
```

"Man" appears `r man_count` times in the text and ranks `r man_rank` among content words once restored. The top 10 words are:

```{r top10-manback}
top10_table(full_sum_table, man_back)
```

### With the onix list excluded entirely

Excluding the onix list reveals additional words suppressed in the original analysis.

```{r cloud-noonix}
#| fig-cap: "Word cloud with the onix stopword list excluded. Additional contextually meaningful words become visible, including 'group', the most frequent word in the volume."
make_word_cloud(table = full_sum_table, stopword_col = no_onix)
```

```{r top-word-noonix}
#| include: false
top_word_noonix <- full_sum_table %>%
  filter(!no_onix) %>%
  slice_head(n = 1) %>%
  pull(word)
```

With the onix list excluded, "`r top_word_noonix`" emerges as the most frequent word, a term with clear contextual relevance to a volume about human social organisation. The top 10 words are:

```{r top10-noonix}
top10_table(full_sum_table, no_onix)
```

### Words removed by the onix list with more than 50 occurrences

The following words were suppressed in the original word cloud by the onix stopword list despite appearing more than 50 times in the main text. Some are genuine function words, but many (including "group", "man", "member", and "area") carry substantive contextual meaning in an analysis of hunter-gatherer social organisation.

```{r onix-removed}
onix_removed <- full_sum_table %>%
  filter(n > 50) %>%
  filter(!no_onix & orig) %>%
  select(word, n) %>%
  arrange(desc(n))

n_onix_removed <- nrow(onix_removed)

dt_table(onix_removed,
         caption    = paste(n_onix_removed, "words suppressed by the onix stopword list with more than 50 occurrences."),
         pageLength = 10)
```

## Gender and Subsistence Term Frequencies

It has also been argued that the *Man the Hunter* volume was broader than commonly portrayed and gave meaningful attention to women's contributions to subsistence. The raw frequencies of key terms can help address this directly.

```{r concept-counts}
key_words <- c("man", "men", "woman", "women",
               "hunt", "hunter", "gather", "gatherer")

concept_counts <- full_sum_table %>%
  filter(word %in% key_words) %>%
  mutate(concept = case_when(
    word %in% c("man", "men")         ~ "man/men",
    word %in% c("woman", "women")     ~ "woman/women",
    word %in% c("hunt", "hunter")     ~ "hunt/hunter",
    word %in% c("gather", "gatherer") ~ "gather/gatherer"
  )) %>%
  group_by(concept) %>%
  summarise(total = sum(n)) %>%
  arrange(desc(total))

concept_counts %>%
  knitr::kable(col.names = c("Concept", "Total occurrences"))
```

```{r concept-inline}
#| include: false

hunt_total        <- concept_counts %>% 
  filter(concept == "hunt/hunter")  %>%
  pull(total)
gather_total      <- concept_counts %>% 
  filter(concept == "gather/gatherer") %>% 
  pull(total)
man_total         <- concept_counts %>% 
  filter(concept == "man/men")        %>% 
  pull(total)
woman_total       <- concept_counts %>% 
  filter(concept == "woman/women")    %>%
  pull(total)

hunt_gather_ratio <- round(hunt_total / gather_total, 1)
man_woman_ratio   <- round(man_total  / woman_total,  1)
```

Hunt/hunting is mentioned `r hunt_gather_ratio` times as often as gather/gathering (`r hunt_total` vs. `r gather_total`) and man/men is mentioned `r man_woman_ratio` times as often as woman/women (`r man_total` vs. `r woman_total`).

## Contextual Analysis: Hunter-Gatherer as a Compound

The frequencies above require an important qualification. "Hunter-gatherer" is a compound term used throughout the volume, and the tokenization process splits it into two separate tokens. This inflates the apparent frequency of both "hunter" and "gatherer" relative to their independent usage. The analysis below examines how often each term appears alongside the other.

```{r context-setup}
words_nearby <- bag_of_words %>%
  filter(!word %in% no_onix_stopwords$word) %>%
  mutate(
    next_word = lead(word),
    pre_word  = lag(word)
  ) %>%
  filter(str_detect(word, "hunt|gath")) %>%
  select(word_orig, pre_word, word, next_word) %>%
  mutate(
    word_s = case_when(
      str_detect(word,      "hunt") ~ "hunt",
      str_detect(word,      "gath") ~ "gather",
      TRUE                          ~ word
    ),
    pre_word_s = case_when(
      str_detect(pre_word,  "hunt") ~ "hunt",
      str_detect(pre_word,  "gath") ~ "gather",
      TRUE                          ~ pre_word
    ),
    next_word_s = case_when(
      str_detect(next_word, "hunt") ~ "hunt",
      str_detect(next_word, "gath") ~ "gather",
      TRUE                          ~ next_word
    )
  )
```

### What precedes "gather"?

```{r gath-pre}
tog_gath <- words_nearby %>%
  filter(str_starts(word_s, "gath")) %>%
  count(pre_word_s, word_s, sort = TRUE) %>%
  mutate(percent = round(100 * n / sum(n), 1))

dt_table(tog_gath,
         caption    = "Words immediately preceding gather/gatherer/gathering.",
         pageLength = 10)
```

### What follows "hunt"?

```{r hunt-post}
tog_hunt_post <- words_nearby %>%
  filter(str_starts(word_s, "hunt")) %>%
  count(word_s, next_word_s, sort = TRUE) %>%
  mutate(percent = round(100 * n / sum(n), 1))

dt_table(tog_hunt_post,
         caption    = "Words immediately following hunt/hunter/hunting.",
         pageLength = 10)
```

### What precedes "hunt"?

```{r hunt-pre}
tog_hunt_pre <- words_nearby %>%
  filter(str_starts(word_s, "hunt")) %>%
  count(pre_word_s, word_s, sort = TRUE) %>%
  mutate(percent = round(100 * n / sum(n), 1))

dt_table(tog_hunt_pre,
         caption    = "Words immediately preceding hunt/hunter/hunting.",
         pageLength = 10)
```

```{r context-inline}
#| include: false
gath_total       <- sum(tog_gath$n)
gath_pct_hunter  <- tog_gath %>% filter(pre_word_s == "hunt") %>% pull(percent)
hunt_total_ctx   <- sum(tog_hunt_post$n)
hunt_pct_gather  <- tog_hunt_post %>% filter(next_word_s == "gather") %>% pull(percent)
top_pre_hunt     <- tog_hunt_pre %>% slice_head(n = 1) %>% pull(pre_word_s)
top_pre_hunt_pct <- tog_hunt_pre %>% slice_head(n = 1) %>% pull(percent)
```

Approximately `r gath_pct_hunter`% of the `r gath_total` occurrences of "gather" are immediately preceded by "hunt", confirming that the majority of gather/gatherer instances are part of the compound "hunter-gatherer" rather than independent references to gathering as a subsistence activity. By contrast, hunt/hunting is followed by a gather variant in approximately `r hunt_pct_gather`% of cases, meaning around three quarters of hunting references stand independently of the compound. The most common word preceding "hunt" is "`r top_pre_hunt`", accounting for `r top_pre_hunt_pct`% of occurrences.

## Appendix: Survey of Stopword Lists

To assess how unusual it is for a stopword list to contain "man" or "men", we examined 68 English stopword lists compiled by [igorbrigadir/stopwords](https://github.com/igorbrigadir/stopwords), a comprehensive repository of stopword lists from NLP tools, search engines, and text analysis libraries.

```{r stopword-survey}
if (!dir.exists("stopwords-master")) {
  download.file(
    "https://github.com/igorbrigadir/stopwords/archive/refs/heads/master.zip",
    destfile = "stopwords-master.zip"
  )
  unzip("stopwords-master.zip")
}

sw_files <- list.files("stopwords-master/en/", pattern = "\\.txt$",
                        full.names = TRUE)

sw_results <- map_df(sw_files, function(f) {
  words <- readLines(f, warn = FALSE) %>% tolower() %>% trimws()
  tibble(
    file      = basename(f),
    n_words   = length(words),
    has_man   = "man"   %in% words,
    has_men   = "men"   %in% words,
    has_woman = "woman" %in% words,
    has_women = "women" %in% words
  )
}) %>%
  left_join(sw_details <- read_csv("stopwords-master/en_stopwords.csv"), by = "file" )

n_lists <- nrow(sw_results)
n_with_man <- sum(sw_results$has_man)


```

```{r stopword-summary}
tibble(
  Term                    = c("man", "men", "woman", "women"),
  `Lists containing term` = c(sum(sw_results$has_man),
                               sum(sw_results$has_men),
                               sum(sw_results$has_woman),
                               sum(sw_results$has_women)),
  `Total lists`           = n_lists,
  `Percent`               = round(100 * c(sum(sw_results$has_man),
                                           sum(sw_results$has_men),
                                           sum(sw_results$has_woman),
                                           sum(sw_results$has_women)) / n_lists, 1)
) %>%
  knitr::kable(
    caption = paste("Frequency of gendered terms across", n_lists, "English stopword lists.")
  )
```

No stopword list contains "woman" or "women". `r n_with_man` lists contain "man" and ``` r``sum(sw_results$has_men) ``` contain "men". These 4 can be seen below. Of them, several seem to have been based on the onix list.

```{r stopword-detail}

sw_results %>%
  filter(has_man) %>%
  select(name, n_words, source_url , date) %>%
  knitr::kable(
    col.names = c("List", "Words", "link", "date"),
    caption   = "Stopword lists containing 'man', none of which contain 'woman'."
  )
```

The onix list was developed by Lextek International in the early 1990s as part of a commercial text retrieval toolkit. Its inclusion of "man" reflects an assumption common in information retrieval of that era: that "man" usually functions as a generic synonym for "human" and is thus uninformative as a search term, while "woman" is a specific context word. This assumption is an example of the inherent bias and ambiguity in English, but also serves as a reminder that careful reading of stopword lists should be a part of any textual analysis.
